Robust inference of neutron-star parameters from thermonuclear burst lightcurves

### 2018-09-23 doc/ms.tex
Fixed bug in fluence method; force inclination to degrees for prior

\title{Thermonuclear burst model-observation comparisons with {\sc concord}}

@@

\author[0000-0002-6558-5121]{Duncan K. Galloway}
\affil{School of Physics \& Astronomy \\
Monash University \\
Clayton, VIC 3800, Australia}

@@

Here we describe a few different approaches that have been used to deduce system parameters from observations of thermonuclear bursts.

@@

Given an estimate of $Q_{\rm nuc}$, the average hydrogen fraction $\langle X\rangle$ in the fuel layer at ignition has been estimated to date using the relation $Q_{\rm nuc}=1.6 + 4\langle X\rangle$~MeV/nucleon \cite[e.g.][and references therein]{gal03d}. This expression includes $\approx35$\% losses
attributed to neutrino emission \cite[]{fuji87}. However, the 35\% value applies only to $\beta$-decays in the much more extensive {\it rp}-process burning chain, and recent {\sc kepler} simulations indicate instead an approximation of $Q{\rm nuc}=0.96+6.6\langle X\rangle$ (Goodwin et al. 2018, in preparation). [{\it comment on improved quadratic expression}]
%
Thus, by substituting in equation \ref{qnuc} we can estimate 
\begin{equation}
\langle X\rangle 
% In [42]: c.c**2/(1.*u.MeV/c.m_p).to("m**2/s**2")/6.6
% Out[42]: <Quantity 142.16243658783114>
% = \frac{c^2z/\alpha}{6.6\ {\rm Mev/nucleon}}\frac{\xi_b}{\xi_p} - 0.145
= z\frac{142}{\alpha}\frac{\xi_b}{\xi_p} - 0.145 \label{xbar}
\end{equation}
[{\it comment on the degree of error introduced by ignoring anisotropy etc.}] 
%
Clearly this expression is only applicable for $\alpha$-values up to some limit, which we calculate as
\begin{equation}
%In [43]: c.c**2/(1.*u.MeV/c.m_p).to("m**2/s**2")/6.6/0.145
%Out[43]: <Quantity 980.4305971574562>
\alpha \leq 980\, z \frac{\xi_b}{\xi_p}
\end{equation}
Interestingly, this limit is $\approx70$\% larger than the equivalent value derived for the old expression for $Q_{\rm nuc}$. This result suggests that larger observed values of $\alpha$ may be accommodated without resorting to explanations including incomplete burning of burst fuel \cite[e.g.][]{bcatalog}.

The hydrogen fraction at ignition $\langle X\rangle$ and the burst recurrence time may then be used to estimate the hydrogen fraction in the fuel, $X_0$. The H-fraction at the base of the fuel layer is reduced steadily by $\beta$-limited hot CNO burning, and will completely exhaust the accreted hydrogen in a time \cite[]{lampe16}:
\begin{equation}
t_{\rm CNO} =9.8\left(\frac{X_0}{0.7}\right) \left(\frac{Z_{\rm CNO}}{0.02}\right)^{-1}
\end{equation}
The hydrogen fraction at the base will thus be $X_0(1-\Delta t/t_{\rm CNO})$, [{\it should this expression include time dilation?}] so that the average H-fraction in the layer will be $\langle X \rangle = X_0(1-0.5\Delta t/t_{\rm CNO})$. We can combine this expression with equation \ref{xbar} to give
\begin{eqnarray}
X_0 & = & \langle X\rangle +0.35\left(\frac{\Delta t}{9.8\ {\rm hr}}\right)
                            \left(\frac{Z_{\rm CNO}}{0.02}\right) \nonumber \\
    & = & z\frac{142}{\alpha}\frac{\xi_b}{\xi_p} - 0.145 
                           +\left(\frac{\Delta t}{28\ {\rm hr}}\right)
                            \left(\frac{Z_{\rm CNO}}{0.02}\right)
\end{eqnarray}
Clearly this estimate of $X_0$ will incorporate systematic errors arising from the assumed values of $z$, the inclination, and $Z_{\rm CNO}$. We explore the typical range of these errors in 
% the next section.
\S\ref{sec:simulations}

@@

>>>>

% \title{Thermonuclear burst model-observation comparisons with {\sc concord}}
\title{Robust inference of neutron-star parameters from thermonuclear burst lightcurves}

@@

\author[0000-0002-6558-5121]{Duncan K. Galloway}

\author{Zac Johnston}
\author{Adelle Goodwin}

\affil{School of Physics \& Astronomy \\
Monash University \\
Clayton, VIC 3800, Australia}

@@

Here we describe three different approaches that have been used to deduce system parameters from observations of thermonuclear bursts.

@@

Given an estimate of $Q_{\rm nuc}$, the average hydrogen fraction $\langle X\rangle$ in the fuel layer at ignition has been estimated to date using the relation $Q_{\rm nuc}=1.6 + 4\langle X\rangle$~MeV/nucleon \cite[e.g.][and references therein]{gal03d}. This expression includes $\approx35$\% losses
attributed to neutrino emission \cite[]{fuji87}. However, the 35\% value applies only to $\beta$-decays in the much more extensive {\it rp}-process burning chain, and recent {\sc kepler} simulations indicate instead an approximation of $Q{\rm nuc}=0.96+6.6\langle X\rangle$ (Goodwin et al. 2018, in preparation). [{\it comment on improved quadratic expression}]
%
Thus, by substituting in equation \ref{qnuc} we can estimate 
\begin{equation}
\langle X\rangle 
% In [42]: c.c**2/(6.6*u.MeV/c.m_p).to("m**2/s**2")
% Out[42]: <Quantity 142.16243658783114>
% = \frac{c^2z/\alpha}{6.6\ {\rm Mev/nucleon}}\frac{\xi_b}{\xi_p} - 0.145
= z\frac{142}{\alpha}\frac{\xi_b}{\xi_p} - 0.145 \label{xbar}
\end{equation}
[{\it comment on the degree of error introduced by ignoring anisotropy etc.}] 
%
Clearly this expression is only applicable for $\alpha$-values up to some limit, which we calculate as
\begin{equation}
%In [43]: c.c**2/(1.*u.MeV/c.m_p).to("m**2/s**2")/6.6/0.145
%Out[43]: <Quantity 980.4305971574562>
\alpha \leq 980\, z \frac{\xi_b}{\xi_p}
\end{equation}
Interestingly, this limit is $\approx70$\% larger than the equivalent value derived for the old expression for $Q_{\rm nuc}$. This result suggests that larger observed values of $\alpha$ may be accommodated without resorting to explanations including incomplete burning of burst fuel \cite[e.g.][]{bcatalog}.

% upper limit on xbar
Conversely, for low values of $\alpha$, we may apply the constraint that $\langle X\rangle \lesssim 0.77$, corresponding to the expected maximum possible for accreted fuel with primordial abundances.

The hydrogen fraction at ignition $\langle X\rangle$ and the burst recurrence time may then be used to estimate the hydrogen fraction in the fuel, $X_0$. The H-fraction at the base of the fuel layer is reduced steadily by $\beta$-limited hot CNO burning, and will completely exhaust the accreted hydrogen in a time \cite[]{lampe16}:
\begin{equation}
t_{\rm CNO} =9.8\left(\frac{X_0}{0.7}\right) \left(\frac{Z_{\rm CNO}}{0.02}\right)^{-1}
\end{equation}
measured in the neutron star frame.
The hydrogen fraction at the base will thus be $X_0[1-\Delta t/(1+z)t_{\rm CNO}]$, 
% [{\it should this expression include time dilation?}] -- yes
so that the average H-fraction in the layer will be $\langle X \rangle = X_0[1-0.5\Delta t/(1+z)t_{\rm CNO}]$. We can combine this expression with equation \ref{xbar} to give
\begin{eqnarray}
X_0 & = & \langle X\rangle +0.35\left[\frac{\Delta t}{(1+z)9.8\ {\rm hr}}\right]
                            \left(\frac{Z_{\rm CNO}}{0.02}\right) \nonumber \\
    & = & z\frac{142}{\alpha}\frac{\xi_b}{\xi_p} - 0.145 
                           +\left[\frac{\Delta t}{(1+z)28\ {\rm hr}}\right]
                            \left(\frac{Z_{\rm CNO}}{0.02}\right)
\end{eqnarray}
Clearly this estimate of $X_0$ will incorporate systematic errors arising from the assumed values of $z$, the inclination, and $Z_{\rm CNO}$. We explore the typical range of these errors in 
% the next section.
\S\ref{sec:simulations}

\begin{itemize}
\item Apply the method to the reference bursts
\end{itemize}

@@

\begin{itemize}
\item Describe Zac's method
\item Describe my method (via multinest)
\end{itemize}
